Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-01 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: parspec review section

BHs in EdGB gravity have scalar hair, to which we can associate a (monopole) scalar charge $Q$,
related to the asymptotic $r^{-1}$ fall-off of the scalar field.
%
This charge is not an independent parameter, and depends on the BH's mass and spin, %~\cite{Berti:2018cxi},
thus being a ``secondary hair''~\cite{Coleman:1991ku,Herdeiro:2015waa}.
%
Since the scalar field is coupled to a curvature scalar, $Q$ is larger (smaller)
the smaller (larger) the BH's mass is.

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar dipole radiation (see e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GWs
observations of compact binaries where at least one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the neutron star-black hole binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm GB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm GB}$ from a selection of black hole binaries (BBH)
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.

@@

More specifically, we follow Refs.~\cite{Cano:2020cao,Cano:2021myl} and work up
to dimension-eight operators ($n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu}\,,
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}}\,,
\label{eq:s_8}
\end{align}
\end{subequations}
%
where ${\cal C} = R_{\mu\nu\rho\sigma} R^{\mu\nu\rho\sigma}$,
$\tilde{\cal C} = R_{\mu\nu\gamma\delta} \epsilon^{\mu\nu}{}_{\rho\sigma} R^{\rho\sigma\gamma\delta}$,
and where $\lambda_{\rm o, e}$ and $\varepsilon_{i}$ ($i=1,2,3$) are dimensionless parameters.
%
Due to the large number of free parameters in this theory, we will work only with a subset of the theory's
parameter space.
%
In particular, we will consider dimension-six and dimension-eight operators separately.
%
In the dimension-six case we further assume that $\lame = \lamo = 1$, leaving us with $\ell = \ell_{\rm cEFT}$
as our single free parameter.
%
In the dimension-eight case, we will set $\varepsilon_{1} = 1$ and $\varepsilon_{2} = \varepsilon_{3} = 0$,
as done in Ref.~\cite{Sennett:2019bpc}.
%
This leaves us with $\ell = \ell_{\rm qEFT}$ as our single free parameter.

@@

In the ParSpec expansion this is accomplished by performing a bivariate
expansion of Eq.~\eqref{eq:general_deviation} in the detector-frame BH's mass
$M$ and spin $\chi$, namely
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{k} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{k} \, \left( 1 + \gamma \delta \omega^{(n)}_{k} \right), \\
\tau_{k}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{k}   \, \left( 1 + \gamma \delta \tau^{(n)}_{k} \right),
\end{align}
\end{subequations}
%
where $\omega_{k}^{n}$ and $\tau_{k}^{n}$ are dimensionless coefficients of the
expansion in spin for Kerr QNMs;
%
$\delta\omega_{k}^{(n)}$ and $\delta\tau_{k}^{(n)}$ are source-independent
dimensionless coefficients that characterize the corrections to the Kerr QNM at
each spin-order;
%
and finally $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on the beyond-GR modification is encapsulated in the dimensionless
parameter $\gamma$ given by,
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on the parameter $\ell$, with dimensions of length (and sets the
lengthscale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
included to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\ell$ dimensionless by using the other
lenghtscale associated to the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$, which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}.

In principle, a modification to GR would also affect $M$ and $\chi$ and the expansion should be written
in terms of non-GR mass and spin, say $\bar{M}$ and $\bar{\chi}$. If we assume that the beyond-GR corrections
are included perturbatively, the modifications to the BH mass and spin can be absorbed into the deviations
parameters $\delta\omega$ and $\delta\tau$.
%
This means we can identify the $M$ and $\chi$ with their \emph{corresponding GR values}.
%
\hs{TODO: say that we check a posteriori if our PE runs agree with this.}

We remark that in the GR limit ($\gamma \to 0$)
% where $\delta\omega_{k}^{(n)} = \delta\tau_{k}^{(n)} = 0$ $\forall\, n$,
the series~\eqref{eq:kerr_expansion} truncated at $N_{\rm max} = 4$, reproduces with 1\% accuracy
the Kerr QNMs for spins $\chi \lesssim 0.7$.
%
The values of the fitting coefficients $\omega_{k}^{(n)}$ and $\tau_{k}^{(n)}$
can be found in Ref.~\cite{Maselli:2019mjd}, Table I.

@@

As we have seen in Sec.~\ref{sec:review_theories}, the QNM of slowly-rotating
BHs is modified gravity theories come in two families depending on their
transformation upon a parity transformation: axial and polar. Which is the one
we use to match against Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work with a chosen theory and perform a
translation between the metric perturbations $h_{\mu\nu}$ in the
Regger-Wheeler-Zerilli gauge~\cite{Regge:1957td,Zerilli:1970se} and connect it with the transverse-traceless gauge
used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Sec. xxx),
%
In GR, both axial and polar QNMs are the same (a property known as \emph{isospectrality}) and therefore which
QNM we use to model the ringdown makes no difference.
%
In beyond-GR theories, isospectrality is in general broken (see the discussion
in Ref.~\cite{Hui:2021cpm} for a counterexample).
%
Thus, how axial and polar gravitational QNMs appear in the GW signal has to be
answered on theory-by-theory basis.
%
This is outside the scope of this paper and here we take the more pragmatic
approach of simply choosing the \emph{least damped gravitational mode between
the two parities.}
%
The justification is this is the mode (if excited during the merger) which is
the most likely to appear in the signal.

@@

%%%%%%%%%%%%%%%%%
\section{Parameter Inference}
\input{sec_pe}

>>>>

BHs in EdGB gravity have scalar hair, to which we can associate a (monopole) scalar charge $Q$,
related to the asymptotic $r^{-1}$ fall-off of the scalar field.
%
This charge is not an independent parameter, and depends on the BH's mass and spin, %~\cite{Berti:2018cxi},
thus being a ``secondary hair''~\cite{Coleman:1991ku,Herdeiro:2015waa}.
%
Since the scalar field is sourced by a curvature scalar, $Q$ is larger (smaller)
the smaller (larger) the BH's mass is.

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar dipole radiation (see e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GW
observations of compact binaries where at least one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the neutron star-black hole binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm GB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm GB}$ from a selection of black hole binaries (BBH)
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.

@@

More specifically, we follow the notation of
Refs.~\cite{Cano:2020cao,Cano:2021myl} and work up to dimension-eight operators
($n=4$),
%
\begin{subequations}
\label{eq:action_mods}
\begin{align}
    L^{(6)} &= \lame R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} R_{\gamma \delta}{}^{\mu\nu}
    + \lamo R_{\mu\nu}{}^{\rho\sigma} R_{\rho\sigma}{}^{\gamma\delta} \tilde{R}_{\gamma \delta}{}^{\mu\nu}\,,
    \label{eq:s_6}
    \\
    L^{(8)} &= \varepsilon_{1} \mathcal{C}^2
    + \varepsilon_{2} \tilde{\mathcal{C}}^{2}
    + \varepsilon_{3} \mathcal{C} \tilde{\mathcal{C}}\,,
\label{eq:s_8}
\end{align}
\end{subequations}
%
where ${\cal C} = R_{\mu\nu\rho\sigma} R^{\mu\nu\rho\sigma}$,
$\tilde{\cal C} = R_{\mu\nu\gamma\delta} \epsilon^{\mu\nu}{}_{\rho\sigma} R^{\rho\sigma\gamma\delta}$,
and where $\lambda_{\rm o, e}$ and $\varepsilon_{i}$ ($i=1,2,3$) are dimensionless parameters.
%
Due to the large number of free parameters in this theory, we will focus on a subset of the theory's
parameter space.
%
In particular, we will consider dimension-six and dimension-eight operators separately.
%
In the dimension-six case we further assume that $\lame = \lamo = 1$, leaving us with $\ell = \ell_{\rm cEFT}$
as our single free parameter.
%
In the dimension-eight case, we will set $\varepsilon_{1} = 1$ and $\varepsilon_{2} = \varepsilon_{3} = 0$,
as done in Ref.~\cite{Sennett:2019bpc}.
%
This leaves us with $\ell = \ell_{\rm qEFT}$ as our single free parameter.

@@

In the ParSpec expansion this is accomplished by performing a bivariate
expansion of Eq.~\eqref{eq:general_deviation} in the detector-frame BH's mass
$M$ and spin $\chi$, namely
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{k} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{k} \, \left( 1 + \gamma \delta \omega^{(n)}_{k} \right), \\
\tau_{k}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{k}   \, \left( 1 + \gamma \delta \tau^{(n)}_{k} \right),
\end{align}
\end{subequations}
%
where $\omega_{k}^{n}$ and $\tau_{k}^{n}$ are dimensionless coefficients of the
expansion in spin for Kerr QNMs;
%
$\delta\omega_{k}^{(n)}$ and $\delta\tau_{k}^{(n)}$ are source-independent
dimensionless coefficients that characterize the corrections to the Kerr QNM at
each spin-order;
%
and finally $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on a given modified gravity theory is encapsulated in the
dimensionless parameter $\gamma$, given by
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on a parameter $\ell$, with dimensions of length (and sets the
lengthscale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\ell$ dimensionless by using the other
lenghtscale associated to the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$, which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}.

In principle, a modification to GR would also affect $M$ and $\chi$ and the
expansion should be written in terms of non-GR mass and spin, say $\bar{M}$ and
$\bar{\chi}$. If we assume that the beyond-GR corrections are included perturbatively
(as is the case for all theories described in Sec.~\ref{sec:review_theories}),
the modifications to the BH mass and spin can be absorbed into the deviations
parameters $\delta\omega$ and $\delta\tau$.
%
This means we can identify the $M$ and $\chi$ with their \emph{corresponding GR values}.
%
We will see that in Sec.~\ref{sec:pe} that this is assumption is indeed
satisfied in our parameter estimation studies (cf.~Fig.~\ref{fig:corner_plot}).

Finally, we remark that in the GR limit ($\gamma \to 0$)
the series~\eqref{eq:kerr_expansion} truncated at $N_{\rm max} = 4$, reproduces with 1\% accuracy
the Kerr QNMs for spins $\chi \lesssim 0.7$.
%
The values of the fitting coefficients $\omega_{k}^{(n)}$ and $\tau_{k}^{(n)}$
can be found in Ref.~\cite{Maselli:2019mjd}, Table I.
%
The fitting coefficients for $N_{\rm max} = 9$, which extend the validity of
the spin-expansion up to $\chi \lesssim 0.99$ where obtained in
Ref.~\cite{Carullo:2021dui}.

@@

As we have seen in Sec.~\ref{sec:review_theories}, the QNM of slowly-rotating
BHs is modified gravity theories come in two families depending on their
transformation upon a parity transformation: axial and polar. Which is the one
we use to match against Eqs.~\eqref{eq:qnm_nongr_iso}?
%
To answer this question one has to work with a chosen theory and perform a
translation between the metric perturbations $h_{\mu\nu}$ in the
Regge-Wheeler-Zerilli gauge~\cite{Regge:1957td,Zerilli:1970se} and connect it with the transverse-traceless gauge
used to described GWs (see e.g.~Ref.~\cite{Maggiore:2018sht}, Sec. xxx),
%
In GR, both axial and polar QNMs are the same (a property known as \emph{isospectrality}) and therefore which
QNM we use to model the ringdown makes no difference.
%
In beyond-GR theories, isospectrality is in general broken (see the discussion
in Ref.~\cite{Hui:2021cpm} for a counterexample).
%
Thus, how axial and polar gravitational QNMs appear in the GW signal has to be
answered on theory-by-theory basis.
%
This is outside the scope of this paper and here we take the more pragmatic
approach of simply choosing the \emph{least damped gravitational mode between
the two parities.}
%
The justification is this is the mode (if excited during the merger) which is
the most likely to appear in the signal.

@@

%%%%%%%%%%%%%%%%%
\section{Parameter Inference}
\label{sec:pe}
\input{sec_pe}
